Analysis and Figures for Gallego et al, 2020

### 2020-09-04 Manuscript/What.lies.ahead.reviewed.tex
replied to comments

Given the sea-surface temperatures and pH values for 2017 (observed) and 2095 (estimated) and taxon-specific logistic regression models, we then evaluated the suitability of habitat for each taxon in the future scenario. For each point in the pH - temperature grid, we calculated a species' probability of presence as the mean of 100 independent draws of the posterior model response. {\color{red} For richness estimations in future scenarios, we used each of the 100 draws of the posteriors in each environmental point as a sample and tallied the number of species detected on each draw. The mean value across the 100 draws was the input for a Wilcoxon test of differences in species richness between the 2017 and 2095 scenarios in each region. We performed the Wilcoxon test globally (total species richness) and on each phylum independently }

@@

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, by 2095 we expect summer in the Hood Canal to be T $>$ 12 and pH $<$ {\color{red} 7.9  \st{8.2}} -- where community 3 is the most likely community -- to occur more than {\color{red} four-fifths \st{two-thirds}} of the time ({\color{red} 83\%  \st{68\%}}). {\color{red} RAMON -- can you see if this 68\% stat is still true, and also if you can incorporate the similar expectations that are commented-out directly below, updating their \% if necessary?} RYAN: It is a tricky sentence as we no longer estimate seasons independently. Cluster 3 is now mode prevalent on pH < 7.8 and Temperature > 12.5; and that condition appears now 83 \% of the time in our simulation.   \\

@@

{\color{red} RAMON - I edited everything above this point. From here down, the text all needs updating to reflect the results in the current figures. }

Diatoms show the steepest richness decline under future conditions (Figure 3 B.2); the probability of occurrence decreases markedly for diverse diatom taxa including both pennate (e.g., {\color{red} \st{\textit{Thalassionema}, \textit{Navicula},}} \textit{Nitzschia} ) and centric (e.g., \textit{Coscinodiscus},  {\color{red} \textit{Stephanodiscus}}) body forms. These declines in diatom richness were more accentuated at lower pH values{\color{red} and higher temperatures}. Other taxa likely to find less-suitable habitat in the future include the dinoflagellates {\color{red}\textit{Karlodinium} and \textit{Gymnodiniaceae} sp.\st{ and red-tide-associated \textit{Chattonella} (haptophyte and raphidophyte, respectively; Hood Canal).} }

Likely winners under future conditions are more widely scattered among higher taxonomic groups. The haptophytes \textit{E. huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium sp.} each finds more suitable habitat in both of our study areas. Among others, \textit{Chaetoceros} (diatom) and {\color{red} many} hydrozoans (Cnidaria) likely increase in San Juan Island, and the potentially fish-killing heterokont flagellate \textit{Pseudochattonella} increases in {\color{red} both study areas}. See Supporting Information for a complete list of taxon-specific projections.

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at richness variation within phyla; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal(taxon richness  by year, Wilcoxon p $=$ 0.7). Diatoms, however, show consistent and significant declines in richness in that region (Wilcoxon p $< 10^{-16}$). }

>>>>

Given the sea-surface temperatures and pH values for 2017 (observed) and 2095 (estimated) and taxon-specific logistic regression models, we then evaluated the suitability of habitat for each taxon in the future scenario. For each point in the pH - temperature grid, we calculated a species' probability of presence as the mean of 100 independent draws of the posterior model response. For richness estimations in future scenarios, we used each of the 100 draws of the posteriors in each environmental point as a sample and tallied the number of species detected on each draw. The mean value across the 100 draws was the input for a Wilcoxon test of differences in species richness between the 2017 and 2095 scenarios in each region. We performed the Wilcoxon test globally (total species richness) and on each phylum independently.

@@

These communities were distinguished by changes in the relative abundances of a wide variety of taxa. In the Hood Canal for example, the colder water and higher-pH community (community 1 in Fig. 2C) showed higher eDNA indexes of diatoms \textit{Thalassiosira nordenskioeldii}, \textit{Nitzschia cf. promare}, and \textit{Ditylum brightwellii} -- among other species -- relative to a higher-temperature community (community 3 in Fig. 2C). Communities 2 and 3 had relatively higher values for taxa from a mix of groups including \textit{Emiliana huxleyi}, \textit{Minutocellus polymorphus}, \textit{Micromonas pusilla}, and \textit{Bathycoccus prasinos}. Community 3 (Fig. 2A,B) and similar planktonic communities occupy the spectrum of environmental conditions most likely to be encountered in near-future climate scenarios as temperature rises and pH falls. For example, by 2095 we expect summer in the Hood Canal to be T $>$ 12\degree C and pH $<$ {\color{red} 7.9  \st{8.2}} -- where community 3 is the most likely community -- to occur more than {\color{red} four-fifths \st{two-thirds}} of the time ({\color{red} 83\%  \st{68\%}}). {\color{red} RAMON -- can you see if this 68\% stat is still true, and also if you can incorporate the similar expectations that are commented-out directly below, updating their \% if necessary?} RYAN: It is a tricky sentence as we no longer estimate seasons independently. Cluster 3 is now mode prevalent on pH < 7.8 and Temperature > 12.5; and that condition appears now 83 \% of the time in our simulation.  {\color{blue} RAMON - ah, yes -- forgot about seasons.  Could we just have some analogous thing where we say what fraction of the time we expect each community, at each area, in 2017 vs 2095?.} \\

@@

Diatoms (Bacillariophyta) show the steepest richness decline under future conditions (Figure 3 B.2); the probability of occurrence decreases markedly for diverse diatom taxa including both pennate (e.g., \textit{Nitzschia}) and centric (e.g., \textit{Coscinodiscus}, \textit{Stephanodiscus}) body forms. These declines in diatom richness were more accentuated at lower pH values and higher temperatures. Other taxa likely to find less-suitable habitat in the future include dinoflagellates \textit{Karlodinium} and Gymnodiniaceae sp.

Likely winners under future conditions are more widely scattered among higher taxonomic groups. The haptophytes \textit{E. huxleyi} and \textit{Chrysochromulina} sp. the dinoflagellate \textit{Alexandrium sp.} each finds more suitable habitat in both of our study areas. Among others not shown in Fig 3, \textit{Chaetoceros} (diatom) and many hydrozoans (Cnidaria) likely increase in San Juan Island, and the potentially fish-killing heterokont flagellate \textit{Pseudochattonella} increases in both study areas. See Supporting Information for a complete list of taxon-specific projections.

Given such heterogeneity in projections, gains and losses tend to balance one another out when looking at richness variation within phyla; {\color{red} we find no change in median richness for the year 2095 relative to the present in the Hood Canal (taxon richness by year, Wilcoxon p $=$ 0.7). Diatoms, however, show consistent and significant declines in richness in that region (Wilcoxon p $< 10^{-16}$). } {\color{blue} RAMON - can you clarify here? there should be comparisons for A. 2017 vs 2095 in overall richness, for both HC and SJI.  Then B. same for Diatom richness, dinoflagellate richness, or anything else you find interesting.}

### 2020-09-04 Scripts/Manuscript.Main.Analysis.Rmd
replied to comments

```{r}
distinct(expanded.future,taxa) %>% 
  mutate(draws = split.matrices.by.taxa) %>% 
  summarise(draws =  Reduce('+',draws)) -> richness.by.draw.by.point

richness.by.draw.by.point %>% 
   nest() %>% 
   mutate(data = map (data, function(.x) colMeans(as.matrix(.x)) %>% 
                        as_tibble %>% 
                        rownames_to_column("sim.row"))) %>% 
   unnest()-> richness.by.point

>>>>

```{r}
# distinct(expanded.future,taxa) %>% 
#   mutate(draws = split.matrices.by.taxa) %>% 
#   summarise(draws =  Reduce('+',draws)) -> richness.by.draw.by.point
# 
# richness.by.draw.by.point %>% 
#    nest() %>% 
#    mutate(data = map (data, function(.x) colMeans(as.matrix(.x)) %>% 
#                         as_tibble %>% 
#                         rownames_to_column("sim.row"))) %>% 
#    unnest()-> richness.by.point

#overall change in richness
expanded.future %>% 
  group_by(Year, Temperature, pH, Area) %>% 
  summarise(richness = sum(present)) %>% 
  group_by(Year, Area) %>% 
  summarise(median(richness))
#diatom change in richness
expanded.future %>% 
  filter(phylum == "Bacillariophyta") %>% 
  group_by(Year, Temperature, pH, Area) %>% 
  summarise(richness = sum(present)) %>% 
  group_by(Year, Area) %>% 
  summarise(median(richness))
